Report on the Second Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE2)

### 2015-06-01 WSSSPE2_report/WSSSPE2_review.tex
finished with section 10 - thanks Tom!

\author{Daniel S. Katz$^{(1)}$, Sou-Cheng T. Choi$^{(2)}$, Nancy Wilkins-Diehr$^{(3)}$, Neil Chue Hong$^{(4)}$, 
\\Colin C. Venters$^{(5)}$, James Howison$^{(6)}$, Frank Seinstra$^{(7)}$, Matthew Jones$^{(8)}$
\\Miguel de Val-Borro$^{(9)}$, Richard Littauer$^{(10)}$, more TBD by contributions}
%\author{Gabrielle Allen$^{(11)}$,  \\
%Karen Cranston$^{(4)}$, Manish Parashar$^{(5)}$, David Proctor$^{(6)}$, \\
%Matthew Turk$^{(2)}$,  Miguel de Val-Borro$^{(10)}$}
%
\thanks{{}$^{(1)}$ Computation Institute, University of Chicago \& Argonne National Laboratory, Chicago, IL, USA; \url{dsk@uchicago.edu}}
%
\thanks{{}$^{(2)}$ NORC at the University of Chicago and Illinois Institute of Technology, Chicago, IL, USA; \url{sctchoi@uchicago.edu}}
%
\thanks{{}$^{(3)}$ University of California-San Diego, San Diego, CA, USA; \url{wilkinsn@sdsc.edu}}
%
\thanks{{}$^{(4)}$ Software Sustainability Institute, University of Edinburgh, Edinburgh, UK; \url{N.ChueHong@software.ac.uk}}
%
\thanks{{}$^{(5)}$ University of Huddersfield, School of Computing and Engineering, Huddersfield, UK}
%
\thanks{{}$^{(6)}$ University of Texas at Austin, Austin, TX, USA; \url{jhowison@ischool.utexas.edu}}
%
\thanks{{}$^{(7)}$ Netherlands eScience Centre, Amsterdam, Netherlands \url{F.Seinstra@esciencecenter.nl}}   
%
\thanks{{}$^{(8)}$ National Center for Ecological Analysis and Synthesis, Santa Barbara, CA, USA; \url{jones@nceas.ucsb.edu}}
%
\thanks{{}$^{(9)}$ Princeton University, Princeton, NJ, USA; \url{mdevalbo@astro.princeton.edu}}  
%
\thanks{{}$^{(10)}$ MIT, Boston, MA, USA \& University of Saarland, Germany \url{richard.littauer@gmail.com}}   
%  
%\thanks{{}$^{(4)}$ National Evolutionary Synthesis Center, Durham, NC, USA}
%
%\thanks{{}$^{(5)}$ Rutgers Discovery Informatics Institute, Rutgers University, New Brunswick, NJ, USA}
%
%\thanks{{}$^{(6)}$ International Consortium of Research Staff Associations, Dublin, Ireland}
%
%\thanks{{}$^{(9)}$ Saarland University, Saarbr\"{u}cken, Germany}
%
%\thanks{{}$^{(10)}$ Department of Astrophysical Sciences, Princeton University, Princeton, NJ, USA}
%
%\thanks{{}$^{(11)}$ University of Illinois, Champaign, IL, USA}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Code Testing \& Code Review} \label{sec:code_testing}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\todo{Dan leading writing of this section}

@@

The consensus of the group was that small, concrete actions would have
a greater chance of proper follow-through by mpmebers of the group.
An example of this approach was the Architecture of Open Source
Applications book~\cite{aosa}, containing contributed chapters on open
source software.  
The group contemplated creating an analogous book/web-resource for
testing of scientific software.

Several members of the discussion indicated that while they valued the
concpet software testing in theory, they were unaware how to test
certain kinds of software.  
Or, more importantly, they lacked relevant examples of software
testing that would be suitable for new members to their team.
The group asked if it could produce a 100-line example of how to test
neutron transport (or other specific scientific examples) targeted at
a sophomore? This would be a demonstration of ``How do I do this
right?''

@@

The group referred to Lorena Barba's ``12 Steps to Navier
Stokes~\cite{12steps},'' an online CFD lesson repository created as a
set of iPython notebooks.  \clunenote{I'm not sure of the context
  here, except possibly as an analog for something under discussion.}

@@

The group recognized that we need to teach people how to test short bits of code, we can't wait until we have thousands of lines of code. Greg commented, ``if we can't teach people how to test code until they're developing thousands of lines of code, we can't be surprised when they don't write tests.''

Distinction between 100 lines of input files vs. 100 lines of source code \katznote{to delete? - don't see how it fits}
\clunenote{I no longer recall the context for this statement; does not seem important to me.}

Some lessons on multiphysics software verification can be found in a recent paper~\cite{dubey-FLASH}.

@@

Two important pieces of the barrier are a) picking/implementing testing frameworks (a technical barrier) and b) deciding what are the actual tests that I need to write?  A possible method to solve the latter is to ask students ``how is what you are doing different from typing random keys on the keyboard?'' and then, turn their answer into the concept of tests.  I.e. expand the process of creating code from one that is solely about ``how?'' into one that also include ``why?''.

@@

The group discussed how they learned testing.  One attendee learned from a tutorial on test driven development (TDD) at a Software Developers Best Practices conference. The examples were {\b not} relevant, partly since he was working with Fortran code in his job, so the remainder of his learning was in the form of proving the process could be applied with Fortran.  Another attendee said that most scientists have broken out pieces of software from a larger application and done some checking in an informal way.  The difference in his work is that he now knows about frameworks that let him do that systematically.  He realized he was always building on top of other people’s software, and eventually realized that he should adopt those good software habits.  He also commented that
most of those initial checks were ephemeral.  They live long enough to give the author of the test confidence in code, then the author moves on; the checks  are not handed on and made available to next user as permanent tests.

@@

See the ``13 dwarves'' from Berkeley http://www.eecs.berkeley.edu/Pubs/TechRpts/2006/EECS-2006-183.html \katznote{to delete? - don't see how it fits}
\clunenote{No opinion - someone thought it was a useful reference for the topic.}

\subsection{Testing examples}

The group took some time to define

\katznote{... working from here still...}

For each problem do:
Write your name and email address so we can get back in touch with you
Write a paragraph or two explaining the scientific problem the code addresses
Tell us how big the simplest piece of relevant, interesting code would be (very rough estimate of lines of code)
Give us a point-form list of the test cases you would use

Tom Clune Problem 1<Thomas.L.Clune@nasa.gov>
Runge-Kutta 4th order
O(100 lines of code)
point-form list of interesting tests
%Test substep:      k’ = F(x + k h, t + h)
%F (x) = C; h = 1; k = 1;   x = 0; expect k’ = C
%F (x) = x^2; h = 2; k = ?;  x = 3; expect k’ = 16
%Test that correct sequence of substeps are used:  use unrealistic F
%F(x) =  {1,4,2,5} on nth call;  h = 1 (unused); x0 = 2;  expect x1 = ?(1*1 + 2*4 + 2*2 + 1*5) = 3
<test with 2D/3D>

Tom Clune Problem 2<Thomas.L.Clune@nasa.gov>
Interpolation with unevenly spaced abscissae
O(200 lines of code)
point-form list of interesting tests
Find bracket
%Trivial case:  nodes @ {1,3}; x = 2;  interval = 1
%nodes@{1,3,5};  x = 4; interval = 2
%How to handle edges? nodes@{1,3,5}; x = 3
%If/how to handle nodes that are not increasing?
%If/how to handle duplicate node?
%Find weights
%End pt. - Interval = {1,2} ; x= 1;  w1 = 1
%Other end pt. - Interval = {1,2}; x = 2; w1 = 0
%middle - Interval = {1,2} x = 1.5; w1 = 0.5
%other sizeInterval = {1, 5}; x = 4; w1 = 0.25
%Combine weights:   
%End pt - w1 = 1;   y1 = 1; y2 = 2;  y = 1
%End pt - w1 = 0;  y1 = 1; y2 = 2; y =2
%Middle - w1 = 0.5;   y1 = 1, y2 = 3;  y = 2
Put it together
At this point really checking that steps are called in order
%interval = bracket(x)
%find weights = (x, interval)
%combine  y = sum(w_i, y_i)
A few simple cases expressed as a table with expected values.

Ethan White Problem 1<ethan@weecology.org>
Logistic growth models are used to model the dynamics of populations of individuals both in isolation and when interacting with other species. At each time step the population of the focal species increases or decreases depending on its reproductive rate, competition from other species, and competition with other individuals of its own species. This growth can be either deterministic or stochastic. Stochastic growth involves some noise in the reproductive rate.
25-100 lines
point-form list of interesting tests
%stochasticity == 0
%r = 0, S=1, x_t = x_0
%r >> 0, S=1, x_1000 = K
%S=2, r1 > r2, K1=K2, x1_1000 = K1, x2_1000 = 0
%S=2 r1=r2, k1=k2, x1_100 = x2_100 = analytical equilibrium 
%stochasticity > 0
%K = 0, x_t = 0
%r1 >> r2, x1_1000 = K1, x2_1000 = K2
pass a “stochasticity” function that changes things in a known way, either always returning the same value or with a set of values that has variation, but whose values are deterministic and then test against known outcomes
e

Ethan White Problem 1<ethan@weecology.org>
Neutral model (based on https://github.com/rueuntal/species-associations/)
100 lines
point-form list of interesting tests
%disp_function == no dispersal
%m == 0, self.immigrants == 0 at any t
pass a “stochasticity” function that changes things in a known way, either always returning the same value or with a set of values that has variation, but whose values are deterministic and then test against known outcomes

Karen Cranston <karen.cranston@nescent.org>
existing SWC example: http://software-carpentry.org/v4/setdict/phylogen.html 
Given DNA sequences for a set of species, we can infer a tree that describes the evolutionary history of the species. We assume that the best tree is the one that involves the fewest number of changes in the DNA (the “maximum parsimony” tree). This code will calculate the parsimony score for a given tree and given DNA alignment. http://telliott99.blogspot.com/2010/03/fitch-and-sankoff-algorithms-for.html 
hmmm…
list of tests
DNA sequences are identical; score = 0
TLC:  Use Fitch’s algorithm for parsimony score.  Assuming your languages does not have built in type for Set.   (E.g. python does, but Fortran does not.) 
%Test Set pseudo-Intersection  “^” (empty intersection is union)
%Single nucleotide tests
%{a} ^ {a} = {a}
%{b} ^ {b} = {b}
%{a} ^ {b} = {a,b
%{a,b} ^ {b,c} = {b}
%Multiple nucleotide tests
%({a},{b})^({b},{b}) = ({a,b},{b})
%Node tests:   Node v has tuple of sets (X_i(v)) and integer score S(v)
%N1 = <({a});s=2>  N2 = <({a});s=3>    ->   Nparent = <({a}; s=2+3>
%N1 = <({a},{a,b});s=2>  N2 = <({b},{b}});s=3>    ->   
%Nparent = <({a,b},{b}); s=2+3+1>
Probably should create a table of N1, N2, Nparent and write a generic test that checks each row of the table.
Tree tests:  Assuming already have post-order binary tree traversal
Tree with 2 leaves and one parent node.   Use table from 3.4.3 to check parent status after traversal with scoring rule
Create more complex tree with 3-4 leaves.  Manually determine status of each top node.
Define parsimony score as s-value of top node

Greg Wilson <gvwilson@software-carpentry.org
Invasion percolation is used to study the spread of pollutants in fractured rock. Model the rock as a 2D grid filled with random integers between 0 and K; mark the central cell as “filled”; and then repeatedly select the cell on the boundary with the lowest value (i.e., the most permeable) and fill it. The density of the resulting fractal tells you how quickly the pollutant spreads.
Simple implementation is ~100 lines of Python.
Interesting tests are:
3x3 grid with a single ‘0’ on the boundary (all other cells are ‘1’) - should fill to boundary in a single step, and the cell with ‘0’ should be the one that fills.
5x5 grid with a straight line of 0’s between the center and the boundary (same case as above, but expanded by one layer)
5x5 grid: cells immediately surrounding the center are all ‘0’, all of boundary is ‘1’ except for one cell, which is 0 - all of the interior should fill, then the single cell on the boundary that’s a ‘1’

For example, look at the testing of the invasion percolation code in this sequence of slides:
The problem: http://software-carpentry.org/v4/invperc/intro.html 
The testing: http://software-carpentry.org/v4/invperc/test.html

>>>>

\author{Daniel S. Katz$^{(1)}$, Sou-Cheng T. Choi$^{(2)}$, Nancy Wilkins-Diehr$^{(3)}$, Neil Chue Hong$^{(4)}$, 
\\Colin C. Venters$^{(5)}$, James Howison$^{(6)}$, Frank Seinstra$^{(7)}$, Matthew Jones$^{(8)}$
\\Thomas L. Clune$^{(9)}$, Miguel de Val-Borro$^{(10)}$, Richard Littauer$^{(11)}$,
\\more TBD by contributions}
%\author{Gabrielle Allen$^{(11)}$,  \\
%Karen Cranston$^{(4)}$, Manish Parashar$^{(5)}$, David Proctor$^{(6)}$, \\
%Matthew Turk$^{(2)}$,  Miguel de Val-Borro$^{(10)}$}
%
\thanks{{}$^{(1)}$ Computation Institute, University of Chicago \& Argonne National Laboratory, Chicago, IL, USA; \url{dsk@uchicago.edu}}
%
\thanks{{}$^{(2)}$ NORC at the University of Chicago and Illinois Institute of Technology, Chicago, IL, USA; \url{sctchoi@uchicago.edu}}
%
\thanks{{}$^{(3)}$ University of California-San Diego, San Diego, CA, USA; \url{wilkinsn@sdsc.edu}}
%
\thanks{{}$^{(4)}$ Software Sustainability Institute, University of Edinburgh, Edinburgh, UK; \url{N.ChueHong@software.ac.uk}}
%
\thanks{{}$^{(5)}$ University of Huddersfield, School of Computing and Engineering, Huddersfield, UK}
%
\thanks{{}$^{(6)}$ University of Texas at Austin, Austin, TX, USA; \url{jhowison@ischool.utexas.edu}}
%
\thanks{{}$^{(7)}$ Netherlands eScience Centre, Amsterdam, Netherlands \url{F.Seinstra@esciencecenter.nl}}   
%
\thanks{{}$^{(8)}$ National Center for Ecological Analysis and Synthesis, Santa Barbara, CA, USA; \url{jones@nceas.ucsb.edu}}
%
\thanks{{}$^{(9)}$ NASA Goddard Space Flight Center, Greenbelt, MD, USA; \url{Thomas.L.Clune@nasa.gov}}  
%
\thanks{{}$^{(10)}$ Princeton University, Princeton, NJ, USA; \url{mdevalbo@astro.princeton.edu}}  
%
\thanks{{}$^{(11)}$ MIT, Boston, MA, USA \& University of Saarland, Germany \url{richard.littauer@gmail.com}}   
%  
%\thanks{{}$^{(4)}$ National Evolutionary Synthesis Center, Durham, NC, USA}
%
%\thanks{{}$^{(5)}$ Rutgers Discovery Informatics Institute, Rutgers University, New Brunswick, NJ, USA}
%
%\thanks{{}$^{(6)}$ International Consortium of Research Staff Associations, Dublin, Ireland}
%
%\thanks{{}$^{(9)}$ Saarland University, Saarbr\"{u}cken, Germany}
%
%\thanks{{}$^{(10)}$ Department of Astrophysical Sciences, Princeton University, Princeton, NJ, USA}
%
%\thanks{{}$^{(11)}$ University of Illinois, Champaign, IL, USA}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Code Testing \& Code Review} \label{sec:code_testing}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\todo{Dan leading writing of this section, with help from Tom Clune}

@@

The consensus of the group was that small, concrete actions would have
the greatest chance of proper followthrough by members of the group.
An example of this approach was the Architecture of Open Source
Applications book~\cite{aosa}, containing contributed chapters on open
source software.  
The group contemplated creating an analogous book or web resource for
testing of scientific software.

Several members of the discussion indicated that while they valued the
concept of software testing in theory, they were unaware how to test
certain kinds of software.  
Or, more importantly, they lacked relevant examples of software
testing that would be suitable for new members to their team.
The group asked if it could produce a 100-line example of how to test
neutron transport (or other specific scientific examples) targeted at
a sophomore? This would be a demonstration of ``How do I do this
right?''

@@

%The group referred to Lorena Barba's ``12 Steps to Navier
%Stokes~\cite{12steps},'' an online CFD lesson repository created as a
%set of iPython notebooks.  \clunenote{I'm not sure of the context
%  here, except possibly as an analog for something under discussion.}

@@

The group recognized that we need to teach people how to test short bits of code, we can't wait until we have thousands of lines of code. Greg commented, ``if we can't teach people how to test code until they're developing thousands of lines of code, we can't be surprised when they don't write tests.''
%
%Distinction between 100 lines of input files vs. 100 lines of source code \katznote{to delete? - don't see how it fits}
%\clunenote{I no longer recall the context for this statement; does not seem important to me.}
%
Some lessons on multiphysics software verification can be found in a recent paper~\cite{dubey-FLASH}.

@@

Two important pieces of the barrier are a) picking/implementing testing frameworks (a technical barrier) and b) deciding what are the actual tests that I need to write?  A possible method to solve the latter is to ask students ``how is what you are doing different from typing random keys on the keyboard?'' and then, turn their answer into the concept of tests.  That is to say, expand the process of creating code from one that is solely about ``how?'' into one that also include ``why?''

@@

The group discussed how they learned testing.  One attendee learned from a tutorial on test driven development (TDD) at a Software Developers Best Practices conference. The examples were {\b not} relevant, partly since he was working with Fortran code in his job, so the remainder of his learning was in the form of proving the process could be applied with Fortran.  Another attendee said that most scientists have broken out pieces of software from a larger application and done some checking in an informal way.  The difference in his work is that he now knows about frameworks that let him do that systematically.  He realized he was always building on top of other people’s software, and eventually realized that he should adopt those good software habits.  He also commented that
most of those initial checks were ephemeral.  They live long enough to give the author of the test confidence in the code, then the author moves on; the checks  are not handed on and made available to next user as permanent tests.

@@

The group then decided that a useful action would be to create a set of scientific codes
with associated tests. For each, they suggested:
\begin{enumerate}
\item Write your name and email address so we can get back in touch with you
\item Write a paragraph or two explaining the scientific problem the code addresses
\item Tell us how big the simplest piece of relevant, interesting code would be (very rough estimate of lines of code)
\item Give us a point-form list of the test cases you would use
\end{enumerate}

%See the ``13 dwarves'' from Berkeley http://www.eecs.berkeley.edu/Pubs/TechRpts/2006/EECS-2006-183.html \katznote{to delete? - don't see how it fits}
%\clunenote{No opinion - someone thought it was a useful reference for the topic.}
